### 2018-07-03 submission/manuscript.Rmd
update abstract

@@

**Methods.** We extracted DNA from fecal samples (n=4) using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach. We first assessed error rate, sequence error prevalence, chimera prevalence, correlation between chimera prevalence and number of Operational Taxonomic Units (OTUs), and GC dependent differences by polymerase and number of cycles. Next, differences in the number of OTUs and taxa was examined based on the polymerase and number of cycles used. Additionally, differences in the bacterial community composition by the Bray-Curtis index also was assessed based on polymerase and number of cycles. Finally, Random Forest models were created to test whether the bacterial community was better at classifying polymerases, number of cycles, or individual donor. We also assessed whether the most important taxa in the polymerase and number of cycle Random Forest models also were the most important in the model for individual donors.

**Results.** When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases (eg. at 35 cycles Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Additionally, the median number of OTUs varied by HiFi DNA polymerase used at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). There also was a large range in the number of OTUs amplified by the polymerases at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2^ = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2^ = `r min(pull(r2_table, r2))`).

**Conclusions.** Although the variation in the number of OTUs in fecal samples could be due to certain polymerases capturing the biological variability better than others, this is unlikely to be the main reason for our observed differences. In mock community samples, the strong correlation between chimera prevalence and the number of OTUs suggests that this is the main reason for differences between the polymerases. Ultimately, this variation makes comparison across studies difficult and care should be exercised when choosing the polymerase and number of cycles in 16S rRNA gene sequencing studies.

>>>>

# Read in GC data
gc_data <- read_csv("../data/process/tables/v4_gc_wilcox_test_summary.csv")

# Read in RF model data
dunn_rf_data <- read_csv("../data/process/tables/dunn_rf_model_test_summary.csv")

@@

**Methods.** We extracted DNA from fecal samples (n=4) using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach. We first assessed GC dependent differences, error rate, sequence error prevalence, chimera prevalence, and correlation between chimera prevalence and number of Operational Taxonomic Units (OTUs) by polymerase and number of cycles. Next, differences in the number of OTUs and taxa was examined based on the polymerase and number of cycles used. Additionally, differences in the bacterial community composition by the Bray-Curtis index also was assessed based on polymerase and number of cycles. Finally, Random Forest models were created to test whether the bacterial community was better at classifying polymerases, number of cycles, or individual donor. We also assessed whether the most important taxa in the polymerase and number of cycle Random Forest models also were the most important in the model for individual donors.

**Results.** Although the total counts for specific taxa varied by polymerase, we found no noticeable differences in total counts based on high and low GC content (P-value $\geqslant$ `r round(min(pull(gc_data, p.value)), digits = 2)`). Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2^ = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2^ = `r min(pull(r2_table, r2))`). When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). There also was a large range in the number of OTUs amplified by the polymerases at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases (eg. at 35 cycles Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Additionally, the median number of OTUs varied by HiFi DNA polymerase used at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). Random Forest models were most successful at classifying individual donor samples rather than polymerase or number of cycles used (P-value $\leqslant$ `r format(max(filter(dunn_rf_data, str_detect(var1, "fecal"), str_detect(var2, "fecal")) %>% pull(bh)), scientific = T, digits = 3)`). Additionally, the most important OTUs in the polymerase and number of cycle models were not the most important in the individual donor sample model.

**Conclusions.** Although there are 16S rRNA gene sequencing differences based on polymerase and number of cycles used, they are small with respect to the biological differences between indivdiuals. Additionally, our observations were not able to identify GC-based differences in amplification based on the number of cycles used. Collectively, these results provide evidence that a real biological difference between groups, based on 16S rRNA gene sequencing, should be detectable regardless of polymerase and number of cycles used.

### 2018-07-03 submission/manuscript.tex
update abstract

\section{Error Introduced into 16S rRNA Gene Sequencing Varies by High
Fidelity DNA Polymerase
Used}\label{error-introduced-into-16s-rrna-gene-sequencing-varies-by-high-fidelity-dna-polymerase-used}

@@

\textbf{Background.} It is challenging to compare 16S rRNA gene
sequencing data across studies and one of the reasons for this is due to
error. There are many different places throughout the workflow where
error can be introduced into the pipeline. Here, we focus on studying
how the number of cycles and high fidelity (HiFi) DNA polymerase
introduce error by varying cycle number and polymerase used to amplify
16S rRNA genes in human fecal and mock community samples.

\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
communities (technical replicates n=4) consisting of previously isolated
whole genomes of 8 different bacteria used the same approach. The
analysis initially examined the number of Operational Taxonomic Units
(OTUs) for fecal samples and mock communities. It also assessed
polymerase and number of cycle dependent differences in the Bray-Curtis
index, error rate, sequence error prevalence, chimera prevalence, and
the correlation between chimera prevalence and number of OTUs.

\textbf{Results.} When analyzing fecal samples we observed that the
range in the number of OTUs detected was not consistent between HiFi DNA
polymerases (eg. at 35 cycles Accuprime = 84 - 106 (min - max) versus
Phusion = 84 - 136). Additionally, the median number of OTUs varied by
HiFi DNA polymerase used at 35 cycles (P-value \textless{} 0.0001). When
analyzing mock community samples the variation in the number of OTUs
detected by the polymerases was observable as early as 20 cycles
(P-value = 0.002). There also was a large range in the number of OTUs
amplified by the polymerases at 35 cycles (Accuprime = 15 - 20 versus
Phusion = 14 - 73). Chimera prevalence in mock communities varied by
polymerase with differences being most notable at 35 cycles (Kappa =
5.71\% (median) versus Platinum = 26.62\%) and this variation persisted
after chimera removal using VSEARCH. We also observed positive
correlations between chimera prevalence and the number of OTUs with
Platinum having the highest (R\textsuperscript{2} = 0.974) and Kappa
having the worst (R\textsuperscript{2} = 0.478).

\textbf{Conclusions.} Although the variation in the number of OTUs in
fecal samples could be due to certain polymerases capturing the
biological variability better than others, this is unlikely to be the
main reason for our observed differences. In mock community samples, the
strong correlation between chimera prevalence and the number of OTUs
suggests that this is the main reason for differences between the
polymerases. Ultimately, this variation makes comparison across studies
difficult and care should be exercised when choosing the polymerase and
number of cycles in 16S rRNA gene sequencing studies.

>>>>

\section{Assessing Differences in High Fidelity DNA Polymerase for 16S
rRNA Gene
Sequencing}\label{assessing-differences-in-high-fidelity-dna-polymerase-for-16s-rrna-gene-sequencing}

@@

\textbf{Background.} A typical 16S rRNA gene sequencing workflow can be
divided into preservation, extraction, amplification, and sequencing
steps. At each of these stages error can be introduced that will change
the underlying bacterial community composition results. In this study we
focus on the amplification step's contribution to this overall error. To
accomplish this we assessed 16S rRNA gene sequencing results in human
fecal and mock community samples after using different high fidelity
(HiFi) DNA polymerases and number of amplification cycles.

\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
communities (technical replicates n=4) consisting of previously isolated
whole genomes of 8 different bacteria used the same approach. We first
assessed GC dependent differences, error rate, sequence error
prevalence, chimera prevalence, and correlation between chimera
prevalence and number of Operational Taxonomic Units (OTUs) by
polymerase and number of cycles. Next, differences in the number of OTUs
and taxa was examined based on the polymerase and number of cycles used.
Additionally, differences in the bacterial community composition by the
Bray-Curtis index also was assessed based on polymerase and number of
cycles. Finally, Random Forest models were created to test whether the
bacterial community was better at classifying polymerases, number of
cycles, or individual donor. We also assessed whether the most important
taxa in the polymerase and number of cycle Random Forest models also
were the most important in the model for individual donors.

\textbf{Results.} Although the total counts for specific taxa varied by
polymerase, we found no noticeable differences in total counts based on
high and low GC content (P-value \(\geqslant\) 0.23). Chimera prevalence
in mock communities varied by polymerase with differences being most
notable at 35 cycles (Kappa = 5.71\% (median) versus Platinum = 26.62\%)
and this variation persisted after chimera removal using VSEARCH. We
also observed positive correlations between chimera prevalence and the
number of OTUs with Platinum having the highest (R\textsuperscript{2} =
0.974) and Kappa having the worst (R\textsuperscript{2} = 0.478). When
analyzing mock community samples the variation in the number of OTUs
detected by the polymerases was observable as early as 20 cycles
(P-value = 0.002). There also was a large range in the number of OTUs
amplified by the polymerases at 35 cycles (Accuprime = 15 - 20 versus
Phusion = 14 - 73). When analyzing fecal samples we observed that the
range in the number of OTUs detected was not consistent between HiFi DNA
polymerases (eg. at 35 cycles Accuprime = 84 - 106 (min - max) versus
Phusion = 84 - 136). Additionally, the median number of OTUs varied by
HiFi DNA polymerase used at 35 cycles (P-value \textless{} 0.0001).
Random Forest models were most successful at classifying individual
donor samples rather than polymerase or number of cycles used (P-value
\(\leqslant\) 5.49e-07). Additionally, the most important OTUs in the
polymerase and number of cycle models were not the most important in the
individual donor sample model.

\textbf{Conclusions.} Although there are 16S rRNA gene sequencing
differences based on polymerase and number of cycles used, they are
small with respect to the biological differences between indivdiuals.
Additionally, our observations were not able to identify GC-based
differences in amplification based on the number of cycles used.
Collectively, these results provide evidence that a real biological
difference between groups, based on 16S rRNA gene sequencing, should be
detectable regardless of polymerase and number of cycles used.
